---
title: "Fst_tutorial"
output: html_document
date: "2023-07-01"
---


# Calculating Fst for the Black Death populations

## 1. Set up
Make sure `vcftools` is installed and added to the PATH;
install and load packages;
locate your working directory.

```{r set up}
## Add vcftools to your PATH
Sys.setenv(PATH=paste(Sys.getenv("PATH"), "/Users/osipova/local/Homebrew/bin/", sep=":"))

## Install and load packages
#install.packages("stringr")
library(stringr)

## Locate your work dir
setwd('/Users/osipova/Documents/LabDocs/TA_human_PopGen/Fst_tutorial/')
```

## 2. Prepare input and run vcftools
Now we want to calculate window-based Fst for the samples from the Black Death study.
I prepared a subsample of variants of *chr5* from the paper and put them in `popgenome-vcf/sampled.chr5.blackDeath.vcf` file. In addition, I prepared two files listing the samples in each population: `before.lst` and `after.lst`
To run `vcftools`, we also need to define the window size and the output file prefix.

```{r vcftools}
## Load VCF input
VCF <- 'popgenome-vcf/sampled.chr5.blackDeath.vcf'
WINDOW = 10000
OUT_PREFIX <- 'before_vs_after_Fst_10kb'

## compute window based Fst
system(str_c('vcftools',
             ' --vcf ', VCF,
             ' --weir-fst-pop before.lst ',
             ' --weir-fst-pop after.lst ',
             ' --fst-window-size ', WINDOW,
             ' --out ', OUT_PREFIX))
```

## 3. Results
Let's check out the output! Load the table in R and check the beginning if the table.
```{r check output}
## Load vcftools output
fst <- read.table(paste0(OUT_PREFIX, '.windowed.weir.fst'), header = TRUE)
head(fst)
```

## 4. Plot Fst
Let's now visualize the results. We will plot Fst values across the entire chr5.

```{r visualize}
## Assign colors and size based on significance
fst$COLOR <- with(fst, ifelse(WEIGHTED_FST > 0.6, 'red', 'black'))
fst$SIZE <- with(fst, ifelse(WEIGHTED_FST > 0.6, 19, 20))

## Plot
par(mfrow=c(1,1))
plot(fst$BIN_START,
     fst$WEIGHTED_FST,
     pch = fst$SIZE,
     col = fst$COLOR,
     ylim = c(0, 1),
     xlab = 'position',
     ylab = 'Fst',
     frame.plot = FALSE,
     main = 'before vs after the Black Death')
```

What are the regions having high Fst values?
We can have a look at the positions with:
```{r}
fst[fst$WEIGHTED_FST > 0.6, ]
```

