---
title: "genetic drift simulations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Assumptions:
+ populatiton size stays constant
+ random mating
+ no mutations
+ no migration
+ no selection

### 1. R function to simulate genetic drift
```{r simulation function}
simulate_genetic_drift <- function(population_size, generations, allele_freq) {
  # Initialize population with initial allele frequencies
  population <- rep(0, population_size)
  population[1:round(allele_freq * population_size)] <- 1
  
  # Simulate generations
  for (gen in 1:generations) {
    # Calculate allele frequencies in the current generation
    allele_freq <- sum(population) / population_size
    
    # Plot the data point
    points(gen, allele_freq, pch=20, col="red")
    points(gen, 1 - allele_freq, pch=20, col='blue')
    
    # Sample the next generation based on the current allele frequencies
    offspring <- sample(population, size = population_size, replace = TRUE)
    
    # Update population with the offspring
    population <- offspring
  }
  
  # Return final allele frequency
  #final_allele_freq <- sum(population) / population_size
  #return(final_allele_freq)
}

```

### 2. Genetic drift in a large population
Let's initiate simulation assigning some values to our parameters.

```{r parameters}
# Set the parameters for the simulation
population_size <- 1000
generations <- 500
allele_freq <- 0.5
```


Now let's run the simulation with assigned parameters
``` {r simulation 1, out.width="65%"}
# Initiate the visualization
plot(0, allele_freq, 
     xlim = c(1, generations),
     ylim = c(0, 1),
     main = "genetic drift",
     xlab = 'number of generations',
     ylab = 'allele frquency', 
     frame.plot = FALSE)

# Run the simulation
simulate_genetic_drift(population_size, generations, allele_freq)

# Print the final allele frequency
#print(paste("Final allele frequency:", final_freq))
```


### 3. Genetic drift in a smaller population

Let's now decrease the population size
```{r parameters 2}
population_size <- 100
```

``` {r simulation 2, echo = FALSE, out.width="65%"}
# Initiate the visualization
plot(0, allele_freq, 
     xlim = c(1, generations),
     ylim = c(0, 1),
     main = "genetic drift",
     xlab = 'number of generations',
     ylab = 'allele frquency', 
     frame.plot = FALSE)
# Run the simulation
final_freq <- simulate_genetic_drift(population_size, generations, allele_freq)
```

We can see how fast one allele went to fixation and the other one disappeared! 


### 4. Genetic drift in a tiny population

Let's now decrease the population size
```{r parameters 3}
population_size <- 10
generations <- 15
```

``` {r simulation 3, echo = FALSE, out.width="65%"}
# Initiate the visualization
plot(0, allele_freq, 
     xlim = c(1, generations),
     ylim = c(0, 1),
     main = "genetic drift",
     xlab = 'number of generations',
     ylab = 'allele frquency', 
     frame.plot = FALSE)
# Run the simulation
final_freq <- simulate_genetic_drift(population_size, generations, allele_freq)
```

It took less than 10 generations for one allele to disappear!

### Conclusion:
In smaller populations the effect of genetic drift in larger.