---
title: "Admixture"
output:
  pdf_document: default
  html_document: default
date: "2023-07-01"
---

This tutorial uses a publicly available dataset from the Human HapMap project (release 23, full dataset). We will use the data in PLINK format from https://zzz.bwh.harvard.edu/plink/res.shtml. This is a big dataset of 270 individuals genotyped at 3.96 million SNPs! What a time to be alive, right?



## 1. Downloading the software/programs
We only need two programs: **ADMIXTURE** and **PLINK**. Obviously, ADMIXTURE does the admixture analysis, while we will use PLINK to wrangle the large dataset into a more tractable for the scope of this tutorial (time and computational constraints).

### For Windows/PC
Sadly, ADMIXTURE is not available for Windows/PC, so folks on that operating system will have to login to the Harvard Cannon cluster (then use the Linux options below).

### For Linux
To download ADMIXUTRE for Linux, type or copy the following into the command line:
```
wget https://dalexander.github.io/admixture/binaries/admixture_linux-1.3.0.tar.gz
```
Unpack the tar ball:
```
tar -xvf admixture_linux-1.3.0.tar.gz --strip-components=1
```
Move the executable to the current directory:
```
mv admixture_linux-1.3.0/admixture .
```
To download PLINK for Linux, type or copy the following into the command line:
```
wget https://s3.amazonaws.com/plink1-assets/plink_linux_x86_64_20230116.zip
```
Unzip the file
```
unzip plink_linux_x86_64_20230116.zip -d plink
```
### For Mac
To download ADMIXUTRE for Mac, type or copy the following into the command line:
```
curl -O https://dalexander.github.io/admixture/binaries/admixture_macosx-1.3.0.tar.gz
```
Unpack the tar ball:
```
tar -xvf admixture_macosx-1.3.0.tar.gz --strip-components=1
```
Move the executable to the current directory:
```
mv dist/admixture_macosx-1.3.0/admixture .
```
To download PLINK for Mac, type or copy the following into the command line:
```
curl -O https://s3.amazonaws.com/plink1-assets/plink_mac_20230116.zip
```
Unzip the file
```
unzip plink_mac_20230116.zip -d plink
```



## 2. Downloading the data
The data can be downloaded directly from the command line. Type or copy the following into the command line:
on Linux:
```
wget https://zzz.bwh.harvard.edu/plink/dist/hapmap_r23a.zip
```
on Mac:
```
curl -O https://zzz.bwh.harvard.edu/plink/dist/hapmap_r23a.zip
```
Unzip the data:
```
unzip hapmap_r23a.zip
```


## 3. Downsampling the data
This is a huge dataset (~4 million SNPs), so lets randomly downsample SNPs to produce a dataset that you can analyze in a few minutes on your laptop. Let’s take a random subsample of 10,000 SNPs as a start. We will use the program PLINK to do this. Type the following at the command line:
```
plink/plink --bfile hapmap_r23a --thin-count 10000 --make-bed --out hapmap_r23_subset
```
If you want to try a larger (or smaller) random subsample, just change the number after the `--thin-count` flag. Pretty easy, right?


## 4. Run analysis
Now the fun part! To run ADMIXTURE, we need to give the program only two things: 
1) the SNP data (your downsampled .bed file created with PLINK) and 
2) a number of ancestral populations (k) to estimate ancestry proportions for.
 
Let's go to RStudio now.
```{r set up}
## Locate ADMIXTURE binary and the working dir
setwd('/Users/osipova/Documents/LabDocs/TA_human_PopGen/Admixture_tutorial/')
ADMIXTURE <- '/Users/osipova/Documents/LabDocs/TA_human_PopGen/software/admixture_macosx-1.3.0/admixture '
```

```{r run test admixture}
knitr::opts_chunk$set(eval = FALSE, eval=FALSE)
k <- 2
system(paste0(ADMIXTURE, 
              'hapmap_r23_subset.bed ',
              k))
```


This runs the ADMIXTURE program assuming 5 ancestral populations (k) and will only take about one minute (for ~10,000 SNPs, but will be longer for larger numbers of SNPs). Really we want to explore various values of k. To assess what the best value of k is, you can have ADMIXTURE conduct cross-validation by simply adding the `--cv` flag to the above command.

```{r run test admixture --cv, eval=FALSE}
knitr::opts_chunk$set(eval = FALSE)
k <- 2
system(paste0(ADMIXTURE, 
              '--cv ',
              'hapmap_r23_subset.bed ',
              k))
```

Building on this, we can write a for loop to run the analysis on multiple values of k (1-5) like so:
```{r run, eval=FALSE }
knitr::opts_chunk$set(eval = FALSE)
## run Admixture for k = 1,2,3,4,5
## store log in a file
for (k in 1:5){
  system(paste0(ADMIXTURE,
                '--cv ',
                'hapmap_r23_subset.bed ',
                k,
                ' > ', k, '.log'))
}
```

## 5. Examine and plot results
Load Admixture output
``` {r load output}
runs <- list()

for (k in 1:5){
  runs[[k]] <- read.table(paste0("hapmap_r23_subset.", k, ".Q"))
}
```

Plot runs for all values of k
```{r plot results}
par(mar=c(4,4,4,1)+.1)
par(mfrow = c(4, 1))

for (k in 2:5){
  barplot(t(as.matrix(runs[[k]])),
          col = rainbow(k),
          xlab = paste0('k = ', k),
          ylab = "Ancestry",
          cex.lab=2,
          border=NA)
}
```



### Cross-validation Results
To view the cross-validation results for each value of k, just type this into the command line:
```
grep  CV *log
```
What we see is:
```
1.log:CV error (K=1): 0.43520
2.log:CV error (K=2): 0.38074
3.log:CV error (K=3): 0.35638
4.log:CV error (K=4): 0.36005
5.log:CV error (K=5): 0.36178
```
From which we can see that the cross-validation estimate is lowest for k=3, suggesting that is a reasonable choice for the value of k. Further analysis, perhaps with a larger number of SNPs could thus focus only on a model where k=3.

